---
title: "Week 5 Conditional Logistic Regression"
author: "Wayne Enanoria"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Show R code in output for all R chunks
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

* Purpose: Code for Week 5 Conditional Logistic Regression Video 
* Data: PHW253_Wk6_CaseControlStudy_redacted_notxt.xlsx


# Preliminary

Call in the necessary R packages for this assignment.

```{r, echo=FALSE}
## load packages
## run everytime

## rio needed for import code
library(rio)
## dplyr needed for mutate or recode
library(dplyr)
## janitor needed for crosstable
library(janitor)
## survival needed for clogit
library(survival)
```

# Data Set

* In Weeks 1-3, we used line listing (a dataset with only the cases)
* In Weeks 4-6, we will use outbreak investigation dataset (a matched case-control study)

```{r}
## import
ecoli <- import("~/PHW253_2023/PHW253_Wk6_CaseControlStudy_redacted_notxt.xlsx")

## look at the dataset 
## example: names(dataset)
dim(ecoli)
names(ecoli)

## janitor package clean names changes the column names to lower case & replaces
## the spaces in column names with an underscore
## example: newnamedataset<- dataset  %>% janitor::clean_names()
ecoli1 <- ecoli %>% janitor::clean_names()
```

# Subset the Data Set for the Case-Control Study

The data set contains more records than needed for the matched case-control study. Thus, you will have to subset the data in order to get the study data set.

```{r}
## Select only cases that have matched controls
ecoli2 <- ecoli1 %>% 
  filter(cdcid !="CDC001" & cdcid !="CDC006" & cdcid !="CDC009" & cdcid !="CDC012" & 
           cdcid !="CDC015" & cdcid !="CDC022" & cdcid !="CDC023" & cdcid !="CDC024" & 
           cdcid !="CDC037" & cdcid !="CDC043" & cdcid !="CDC045" & cdcid !="CDC046" & 
           cdcid !="CDC048" & cdcid !="CDC049" & cdcid !="CDC067" & cdcid !="CDC068")   

## Some cases have multiple controls
ecoli2 %>%  filter(cdcid == "CDC047")  %>% tabyl(case)

## Only keep controlletter=A for cdcid=="CDC047"
x <- which(ecoli2$cdcid =="CDC047" & ecoli2$controlletter=="B")   
ecoli3 <- ecoli2[-x, ]    

dim(ecoli3)
```

# Recode Variables

For conditional logistic regression, all food item values should be 0 or 1. Recode other values to missing. You can use factors also but that is beyond the scope of this assignment, so we will simply recode the variables.

```{r}
ecoli3 %>%  tabyl(milk, case)

## milk is the only variable coded as: 0,1,NA
## all other variables have some NAs coded as 99 and/or 3

ecoli3$xrawcd <- ifelse((ecoli3$rawcd==3 | ecoli3$rawcd==99),NA,ecoli3$rawcd)
ecoli3 %>%  tabyl(xrawcd, rawcd)
ecoli3 %>%  tabyl(xrawcd, case)
```


# Create a numeric variable that identifies each pair (case and control pair), also called a statum or group variable.

* This is called strata in the regression code.
* First, select last two characters of the CDCID string/character variable (new variable stratachar)
* Then, convert to a numeric format (new variable called stratanum)

```{r}
ecoli3$stratachar <- substr(ecoli3$cdcid,start=5,stop=6)

ecoli3 %>%  tabyl(stratachar)

ecoli3$stratanum <- as.numeric(ecoli3$stratachar)

ecoli3 %>%  tabyl(stratachar,stratanum)
```

# Conduct the conditional logistic regression model using one exposure and one outcome.
 
```{r}
clogit(case ~ milk + strata(stratanum), data=ecoli3)
## the exponential of the coefficient "exp(coef)" is the odds ratio of 2.0
```


```{r}
## below is coding trick to output exponentiated 95% CIs
## these are the 95% CI of the odds ratio
m <- clogit(case ~ milk + strata(stratanum), data=ecoli3)
summary(m)
## probability of observing a z-statistic value more extreme than
## the one observed "Pr(>|z|)" is the p-value of 0.571.
```

# Additional example: raw cookie dough variable

```{r}
m<-clogit(case ~ xrawcd + strata(stratanum), method="exact", data=ecoli3)
summary(m)
```


```{r}
## if model does not converge, try to change to a non-exact method

m<-clogit(case ~ xrawcd + strata(stratanum), method="approximate", data=ecoli3)
summary(m)

## why does the model (still) not converge? let's look at the data
ecoli3 %>% tabyl(xrawcd, case)
ecoli3 %>% tabyl(xrawcd, case, cdcid)
```

* column A (xrawcd=0 for control & case): 2
* column B (xrawcd=0 for control, 1 for case): 29
* column C (xrawcd=1 for control, 0 for case): 0
* column D (xrawcd=1 for control & case): 4
* dropped observation/missing exposure info: 1

# Note 

In practice, we would then use a different regression model made for small cells called exact logistic regression. R does not have a readily available exact logistic regression package like Stata. (The archived elrm package uses a Markov Chain Monte Carlo algorithm to conduct an approximation for exact logistics regression which we will not use here.)

So for raw cookie dough, the matched Odds Ratio, 95% Confidence Interval, and p-value are provided (obtained from study authors):

* Odds Ratio: 41.3 
* 95% confidence interval (7.37, >999.99) 
* p-value <0.001

Use these values in your assignment so you have the correct OR and 95% CI for the raw cookie dough exposure variable.
 

 

 




 



