---
title: "PHW253 Logistic Regression"
author: "Wayne Enanoria"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Show R code in output for all R chunks
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

* Purpose: Code for Week 4 Logistic Regression
* Data: PHW253Oswegorevised.csv


# Preliminary

We will not need any special R packages for this assignment.

# Data Set

Next, call in the data.

```{r}
oswego <- read.csv("~/PHW253_2023/PHW253Oswegorevised.csv",
                     header=TRUE,
                     na.strings=".")
dim(oswego)
names(oswego)
```

# Data Management for Logistic Regression

The glm() function (generalized linear model) we will use in R requires that the variables of the model be numeric or factor variables. So we will have to do some data manipulation with our data before we can run the R code to conduct logistic regression.

```{r}
# Examine the outcome variable, ill
table(oswego$ill, exclude=NULL)

# Create a numeric variable for Ill
oswego$xill <- ifelse(oswego$ill=="Y", 1, 0)
# Check your code
table(oswego$xill, oswego$ill, exclude=NULL)

# Examine the exposure variable, mashedpotato
table(oswego$mashedpotato, exclude=NULL)

# Create a numeric variable for mashedpotato
oswego$xmashedpot <- NA
oswego$xmashedpot[oswego$mashedpotato=="Y"] <- 1
oswego$xmashedpot[oswego$mashedpotato=="N"] <- 0
table(oswego$xmashedpot, oswego$mashedpotato, exclude=NULL)

# Look at the two-by-two table
table(oswego$mashedpotato, oswego$xill, exclude=NULL)

# Run the logistic regression
rslt2 <- glm(xill~ xmashedpot, data=oswego, family=binomial(link="logit"))
summary(rslt2)

# Estimate the odds ratio
exp(coef(rslt2))
# Only the odds ratio
exp(coef(rslt2)[2])

# Confidence intervals
exp(confint.default(rslt2))
```


