--
title: "PHW253 Descriptive Epidemiology Assignment"
author: "Wayne Enanoria"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Show R code in output for all R chunks
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

* Purpose: Code for Week 3 Descriptive Epi Assignment
* Data: PH W253 Wk4 Assignment line list_redacted.xlxs


# Preliminary

You may have to install new packages for this assignment. 

```{r, include=FALSE}
## load packages
## run everytime
## rio needed for import code
library(rio)

## dplyr needed for mutate or recode
library(dplyr)

## janitor needed for crosstable
library(janitor)

## flextable needed for export formatted table to Word file
library(flextable)

# for dates in R
library(lubridate)

# ggplot2 for epicurves & data visualization
library (ggplot2)

# incidence2 for epicurves of linelist data
library (incidence2)
```

# Data Set

Next, call in the data.

```{r}
outbreak <- import("~/PHW253_2023/PH W253 Descriptive Epi Assignment line list_redacted.xlsx")
names(outbreak)
head(outbreak)
```

# Assignment

You can break this code up into smaller R chunks if you wish.

```{r}
## Descriptive Epi Assignment instructions  
## state "same outbreak MLVA pattern (Pattern A, A2, and A3)"
table(outbreak$MLVA)
table(outbreak$MLVA, exclude=NULL)

## there are 92 observations in the data set
## 75 cases had are MLVA pattern A, 1 case pattern A2, 1 case pattern A3
## other E.coli cases did not have MLVA pattern A, A2, pr A3 and are not a part of this outbreak

## create new dataset where only the statement within parentheses is TRUE is kept
# uses dplyr package 
outbreakA <- outbreak %>% 
             filter(MLVA == "A" |MLVA == "A2" |MLVA == "A3")   
# keep only rows where MLVA is an A, A2,A3 pattern

head(outbreakA)


```

# Dates in R

Now we will handle the dates.

```{r}
## Dates in R
## Step 1: What format is the OnsetDate in?
class(outbreakA$OnsetDate)

## Step 2: use package lubridate to convert to Date format
outbreakA <- outbreakA %>%
             mutate(newOnset = lubridate::ymd(OnsetDate))

## Step 3: confirm that new variable format
class(outbreakA$newOnset)

table(outbreakA$newOnset, exclude=NULL)
## dates range from March to July 2009 & there are missing values
```



# Epidemic Curve by Day

Now we create an epidemic curve.

NOTE: It is difficult to see epi peak on daily curve. The rest of code creates a WEEKLY epi curve in using incidence2 package or ggplot package

## Using incidence2 package, Create an epi_day dataframe


```{r}
epi_day <- incidence(       # create incidence object
  x = outbreakA,             # dataset
  date_index = "newOnset",  # date column
  interval = "day"          # date grouping interval
)

# NOTE: the log states that 7 missing observations were removed from epi_day.
epi_day

# Plot the epi curve
plot(epi_day)

## Notes: 
##   1) epi curve is missing title & label on x-axis & footnote indicating missing data     
##   2) epi curve starts mid-March
##   3) epi curve has grid lines
```
 
```{r}
# Plot the epi curve with labels and formatting

plot(epi_day, xlab = "Date of illness onset") + 
  labs(title = "Figure 1. Epidemic Curve for an outbreak of E. coli O157:H7, March 1 – July 31, 2009", caption="Figure does not include n=7 individuals with missing onset date.") + 
  scale_x_date(limits = c(as.Date("2009-03-01"), NA)) +
  theme_classic()
```


## Using incidence2 package, epidemic curve by week

```{r}
epi_week <- incidence(              # create incidence object
            x = outbreakA,          # dataset
            date_index = "newOnset",  # date column
            interval = 7            # date grouping interval e.g., every 7 days
)
## NOTE: log states that 7 missing observations were removed from epi_week

epi_week
 
# Plot the epi curve
plot(epi_week)

## If above code will not plot use the below code
epi_week <- incidence(              # create incidence object
            x = outbreakA,          # dataset
            date_index = "newOnset",  # date column
            interval = "week"            # date grouping interval e.g., every 7 days
)
## NOTE: log states that 7 missing observations were removed from epi_week

epi_week
 
# Plot the epi curve
plot(epi_week)


```

## Refine the epidemic curve

```{r}
plot(epi_week, xlab = "Date of illness onset") + 
  labs(title = "Figure 1. Epidemic Curve for an outbreak of E. coli O157:H7, March – July, 2009", caption="Figure does not include n=7 individuals with missing onset date.") + 
  theme_classic()
```




## Using ggplot, create a weekly epidemic curve

The below code provides plot with all labels

```{r}
weekly_breaks <- seq.Date(from = as.Date("2009-03-01"),
                          to = as.Date("2009-07-31"),
                          by = "week")

ggplot(data = outbreakA) + 
  # make histogram: specify bin break points above
  geom_histogram(
  # mapping aesthetics
  mapping = aes(x = newOnset),  # date column mapped to x-axis
  # histogram bin breaks
  breaks = weekly_breaks, # histogram bin breaks defined previously
  # bars
  color = "darkblue",     # color of lines around bars
  fill = "lightblue"      # color of fill within bars
  ) + 
    # y-axis
    scale_y_continuous(  breaks = seq(from = 0, to = 14, by = 2),  # specify y-axis increments by 2
                       expand = c(0,0))+             # remove excess y-axis space below 0 (align histogram flush with x-axis)

  # aesthetic themes
  theme_classic() +                # simplify plot background
  theme(
    plot.caption = element_text(hjust = 0)) +       # caption on left side
    # labels including caption
    labs(
    title    = "Figure 1. Epidemic Curve for an outbreak of E. coli O157:H7, March 1 – July 31, 2009" ,
    x        = "Date of illness onset" ,
    y        = "Weekly incident cases" ,
    caption  = "Figure does not include n=7 individuals with missing onset date."
    )

```


