--
title: "Week 2: Descriptive Statistics using Oswego Data Set"
subtitle: "Week2-DescriptiveStatistics.Rmd"
author: "Wayne Enanoria"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Show R code in output for all R chunks
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The purpose of this file is to show how to conduct descriptive statistics using R. 

In this session, we will use data from a church outbreak in a small village of Lycoming, Oswego County, NY, on April 18, 1940. 


## Step 1: Import the data set and conduct descriptive statistics using one variable.

```{r}
# Import the data
oswego <- read.csv("~/PHW253_2022/PHW253Oswegorevised.csv",
                   header=TRUE,
                   na.strings=".")
# look at the number of records (first number) and variables (second number)
dim(oswego)

# Summarize age variable
summary(oswego$age)

# Create a histogram of the age variable
hist(oswego$age, main="Histogram of Age (Years)", xlab="Age (Years)")

# Create a boxplot of the age variable
boxplot(oswego$age, main="Boxplot of Age (Years)")

# Show other quantiles
quantile(oswego$age)
quantile(oswego$age, probs=c(0.25, 0.50,  0.75, 0.90, 0.99))

# look at the mean, median, and range
mean(oswego$age)
median(oswego$age)
range(oswego$age)

# min and max
min(oswego$age)
max(oswego$age)
```

## Step 2: Descriptive Statistics of one continuous variable across one categorical variable

```{r}
# Create a boxplot
boxplot(oswego$age ~ oswego$sex, 
        main="Boxplot of Age (Years) by Gender",
        xlab="Gender",
        ylab="Age (Years)")

tapply(oswego$age, oswego$sex, summary,)
tapply(oswego$age, oswego$sex, quantile, probs=c(0.25, 0.50,  0.75, 0.90, 0.99))

# Recode gender variable
table(oswego$sex, exclude=NULL)

# copy the original variable
oswego$gender <- oswego$sex

# recode the male category
oswego$gender[oswego$sex=="M"] <- "Male"

# recode the female category
oswego$gender[oswego$sex=="F"] <- "Female"

# check your work
table(oswego$sex, oswego$gender, exclude=NULL)

# Now you can re-run previous boxplot with recoded gender variable
boxplot(oswego$age ~ oswego$gender, 
        colors=c("blue", "gray"),
        main="Boxplot of Age (Years) by Gender",
        xlab="Gender",
        ylab="Age (Years)")
```

## Step 3: Crosstable with two categorical variables

```{r}
# For this one, we will use the janitor R package
library(janitor)

# Recreate age variable from Week 1
oswego$agecat <- cut(x=oswego$age,
                     breaks=c(0, 1, 5, 10, 20, 30, 40, 50, 60, 200),
                     labels=c("<1", "1-5", "6-10", "11-20", "21-30","31-40", "41-50", "51-60", "61+"))

oswego %>% tabyl(agecat)

# format the percents properly
oswego %>% tabyl(agecat) %>% adorn_pct_formatting() 

# you can look at two categories at one time
oswego %>% tabyl(agecat, sex)
oswego %>% tabyl(agecat, ill)

# you can look at each gender separately
oswego[oswego$sex=="M",] %>% tabyl(agecat)
oswego[oswego$sex=="F",] %>% tabyl(agecat)


## display counts & percents and include labels
oswego %>%                                    # case linelist / data source
  tabyl(agecat, ill) %>%                      # cross-tabulate counts
  adorn_totals(where = "row") %>%             # add a total row
  adorn_percentages(denominator = "col") %>%  # convert to proportions
  adorn_pct_formatting() %>%                  # convert to percents
  adorn_ns(position = "front") %>%            # display as: "count (percent)"
  adorn_title(                                # adjust titles
    row_name = "Age Category",
    col_name = "Ill")
```

## Step 4: Save result to a MS Word document

```{r}
# Bonus task: export to a MS Word document
library(flextable)

oswego %>%                                    # case linelist
  tabyl(agecat, ill) %>%                      # cross-tabulate counts
  adorn_totals(where = "row") %>%             # add a total row
  adorn_percentages(denominator = "col") %>%  # convert to proportions
  adorn_pct_formatting() %>%                  # convert to percents
  adorn_ns(position = "front") %>%            # display as: "count (percent)"
  adorn_title(                                # adjust titles
    row_name = "Age Category",
    col_name = "Ill",
    placement = "combined") %>% 
  flextable::flextable() %>%                  # convert to image
  flextable::autofit() %>%                    # ensure only one line per row
  flextable::save_as_docx(path = "~/PHW253_2022/tabyl_ageill.docx")   # save as Word document
```

## Step 5: Creating a table using kable()

```{r}
# Usekable to create a table
library(knitr)

table(oswego$agecat)
kable(table(oswego$agecat))
kable(table(oswego$agecat), col.names=c("Age (Years)","Frequency"), align="lc")

table(oswego$agecat, oswego$ill, exclude=NULL)
kable(table(oswego$agecat, oswego$ill, exclude=NULL))
kable(table(oswego$agecat, oswego$ill, exclude=NULL), caption="Frequency of Age (years) by Ill Status")


```

This is the end of the file.


